---
title: 'Reproducible Research: Peer Assessment 1'
author: "Mohsen Najafi Yazdi"
output: html_document
---

This is a report on statistical investigation of a sample [personal activity monitoring device dataset](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) as the first project of Reproducible Research course. This report is publicly avaliable at: https://github.com/MarlexBlank/RepData_PeerAssessment1

## Loading and preprocessing the data

```{r, echo= TRUE}
  dataFile <- unz("activity.zip","activity.csv")
  rawData <- read.csv(dataFile,header = TRUE,sep = ",",stringsAsFactors = FALSE)
  head(rawData)
```

## What is mean total number of steps taken per day?
To answer this question, we first need to calculate the number of steps taken __each day__.

#### Calculating the total number of steps taken each day:
```{r, echo= TRUE}
  perDayData <- split(rawData,rawData$date)
  numDays <- length(perDayData)
  stepsPerDay <- data.frame(dates = names(perDayData) , steps = 0)
  for ( iDay in 1:numDays )
  {
    todayData <- perDayData[iDay]
    todayData <- do.call(rbind.data.frame,todayData)
    stepsPerDay$steps[iDay] <- sum(todayData$steps)
  }
  
```
The variable `stepsPerDay` contains the desired data.

#### Plotting histogram of number of steps taken per day:
```{r, echo= TRUE}
  hist(stepsPerDay$steps , main = "Histogram of number of steps per day", 
       col = "green" , xlab = "Number of steps per day" , breaks = 10)
```

#### Mean and median of number of steps taken per day:
```{r, echo= TRUE}
  stepsMean <- mean(stepsPerDay$steps[!is.na(stepsPerDay$steps)]) 
  print(paste("Mean = ",stepsMean))
  
  stepsMedian <- median(stepsPerDay$steps[!is.na(stepsPerDay$steps)]) 
  print(paste("Median = ",stepsMedian))
```

## What is the average daily activity pattern?
To answer this question, we first need to create a time series for __each 5-min interval__.

#### Creating the desired data:
```{r, echo= TRUE}
  timeSeries <- split(rawData,rawData$interval)
  numIntervals <- length(timeSeries)
  averageStepsPerInterval <- data.frame(interval = names(timeSeries) , avgerageSteps = 0)
  for ( iInterval in 1:numIntervals )
  {
    thisIntervalData <- timeSeries[iInterval]
    thisIntervalData <- do.call(rbind.data.frame,thisIntervalData)
    averageStepsPerInterval$avgerageSteps[iInterval] <- mean(
      thisIntervalData$steps[!is.na(thisIntervalData$steps)])
  }
```
The variable `averageStepsPerInterval` contains the desired data.

#### Ploting the time series:
```{r, echo= TRUE}
  plot(as.character(averageStepsPerInterval$interval),averageStepsPerInterval$avgerageSteps,type = "l", 
       xlab = "Interval" , ylab = "Average # of steps" , main = "Average number of steps")
```

#### Finding the inverval which, on average across all days, has the maximum number of steps:
```{r, echo= TRUE}
  maxIndex = which(averageStepsPerInterval$avgerageSteps == max(averageStepsPerInterval$avgerageSteps))
  print(paste("The interval",as.character(averageStepsPerInterval$interval[maxIndex[1]]),
              " has the maximum average number of steps which is",
              as.character(averageStepsPerInterval$avgerageSteps[maxIndex[1]])))
```

## Imputing missing values
In this section, we deal with missing values and replace them with real values.
#### Finding the total number of missing values:
```{r, echo= TRUE}
  print(paste("The total number of missing values is",sum(is.na(rawData$steps))))
```
#### Strategy for filling the missing values:
I will use the __mean of 5-min intervals across all days__ to fill in the missing values. The reason behind it is that a person most likely has a fixed daily schedule. Hence, the general pattern of a day, on average, could be a good alternative for missing values.

#### Calculating the mean of 5-min interval across days:
```{r, echo= TRUE}
  timeSeries <- split(rawData,rawData$interval)
  numIntervals <- length(timeSeries)
  meanStepsPerInterval <- data.frame(interval = names(timeSeries) , meanSteps = 0)
  for ( iInterval in 1:numIntervals )
  {
    thisIntervalData <- timeSeries[iInterval]
    thisIntervalData <- do.call(rbind.data.frame,thisIntervalData)
    meanStepsPerInterval$meanSteps[iInterval] <- mean(
      thisIntervalData$steps[!is.na(thisIntervalData$steps)])
  }
```
The variable `meanStepsPerInterval` contains the desired data.

#### Replacing missing values with the calculated data:
```{r, echo= TRUE}
  naIndex = which(is.na(rawData$steps) == TRUE)
  numMissingData = length(naIndex)
  cleanData <- rawData
  for ( i in 1:numMissingData )
  {
    cleanData$steps[i] <- meanStepsPerInterval$meanSteps[
      which(meanStepsPerInterval$interval == rawData$interval[i])]
  }
  head(cleanData)
```
The variable `cleanData` contains the desired data.

#### Plotting histogram of clean data, and calculating mean and median as before:
```{r, echo= TRUE}
  perDayData <- split(cleanData,cleanData$date)
  numDays <- length(perDayData)
  stepsPerDay <- data.frame(dates = names(perDayData) , steps = 0)
  for ( iDay in 1:numDays )
  {
    todayData <- perDayData[iDay]
    todayData <- do.call(rbind.data.frame,todayData)
    stepsPerDay$steps[iDay] <- sum(todayData$steps)
  }
  
  hist(stepsPerDay$steps , main = "Histogram of number of steps per day (clean data)", 
       col = "green" , xlab = "Number of steps per day" , breaks = 10)

  stepsMean <- mean(stepsPerDay$steps[!is.na(stepsPerDay$steps)]) 
  print(paste("Mean (clean data) = ",stepsMean))
  
  stepsMedian <- median(stepsPerDay$steps[!is.na(stepsPerDay$steps)]) 
  print(paste("Median (clean data) = ",stepsMedian))
```

#### The impact of imputing missing data on the estimates of the total daily number of steps:
Since I used the __mean of 5-min intervals across all days__, it did not have a significant impact on the distribution of total daily number of steps. The reason is because __total daily number of steps__ is simply the summation over all intervals. Hence, replacing the missing value with the __mean value of that interval__ is basically a realization from the same distribution.

## Are there differences in activity patterns between weekdays and weekends?
#### Creating a new factor for __weekdays__ and __weekend__:
```{r, echo= TRUE}
  cleanData$dayType <- ifelse(weekdays(as.Date(cleanData$date)) == 
      "Sunday" | weekdays(as.Date(cleanData$date)) == "Saturday",
      "weekend","weekday")
  head(cleanData)
```
The variable `cleanData` contains the desired data.

#### Plotting the 5-min interval time series for weekdays and weekends:
```{r, echo= TRUE}
  timeSeries <- split(cleanData,cleanData$interval)
  numIntervals <- length(timeSeries)
  averageStepsPerInterval <- data.frame(interval = names(timeSeries) , weekdaySteps = 0 , weekendSteps = 0)
  for ( iInterval in 1:numIntervals )
  {
    thisIntervalData <- timeSeries[iInterval]
    thisIntervalData <- do.call(rbind.data.frame,thisIntervalData)
    averageStepsPerInterval$weekdaySteps[iInterval] <- mean(
      thisIntervalData$steps[!is.na(thisIntervalData$steps) & thisIntervalData$dayType == "weekday"])
    averageStepsPerInterval$weekendSteps[iInterval] <- mean(
      thisIntervalData$steps[!is.na(thisIntervalData$steps) & thisIntervalData$dayType == "weekend"])
  }
  par(mfrow=c(2,1))
  plot(as.character(averageStepsPerInterval$interval),averageStepsPerInterval$weekdaySteps,type = "l", 
       xlab = "Interval" ,ylab = "", main = "Average number of steps (weekdays)")
  plot(as.character(averageStepsPerInterval$interval),averageStepsPerInterval$weekendSteps,type = "l", 
       xlab = "Interval" ,ylab = "", main = "Average number of steps (weekends)")
```